An Algorithm to Prune Integer Overflow Tests

### 2012-08-30 doc/Manuscript/CGO/paper.tex
Added Raphael's text to the paper.

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass.
This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \url{http://llvm.org/docs/LangRef.html}}
instead of working directly on the original source code. 
This fact brings the advantage of allowing us to rely on
other analysis, such as the range analysis, to check if it is safe to state that
the possible values of an instruction will always be within the range of
the declared type of the instruction. 
If we can state this, it means that it is impossible for an overflow to 
occur in the analyzed instruction, because the range analysis is a conservative analysis; 
in this case, we don't need to insert dynamic checks, and the instrumented program
will be smaller and run faster than it would run without this prunning.
%How our instrumentation works
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after that instructions. 
The LLVM IR has five instructions that may lead to an overflow: 
ADD, SUB, MUL, SHL, and TRUNC.

@@

\noindent
\textbf{Precision.}
Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, we present a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark suite, which is publicly
available~\footnote{\url{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}}.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due to library functions that we cannot
analyze.

@@

\noindent
\paragraph{Dynamic Detection of Integer Overflows: }
Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to detect
integer integer overflows.
Their approach consists in instrumenting every integer operation that might cause
to an overflow, underflow, or data loss.
Different from our method, RICH uses specific features of the x86 architecture
to reduce the instrumentation overhead.
Moreover, no static analysis is used to avoid unnecessary assertions, although
this improvement is cited as future work.

\paragraph{Static Detection of Integer Overflows: }
\noindent
Zhang {\em et al.}~\cite{Zhang09} have designed a static analysis to sanitize
programs against integer overflow based vulnerabilities.
They instrument integer operations in paths from a source to a sink.
In Zhang {\em et al.}'s context, sources are functions that read values from
users, and sinks are memory allocation operations.
Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
about 10\% of the integer operations in the program.
However, they do not use any form of range analysis to limit the number of
checks inserted in the transformed code.
Wang {\em et a.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities. The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software. Our work, and Wang {\em et a.}Õs work are essentially different: they use symbolic execution, whereas we rely on
range analysis.
Furthermore, contrary to them, we do not try to detected statically the
possibility of an integer overflow happening.
Contrary to us, they do not transform the program to prevent or detect such event dynamically.
Still in the field of symbolic execution, Molnar
{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
Linux x86 binaries to find integer overflow bugs.
They prove the existence of bugs by generating test cases for them.

>>>>

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass.
This means our library works on the LLVM IR\footnote{LLVM Intermediate Representation, available at \texttt{http://llvm.org/docs/LangRef.html}}
instead of working directly on the original source code. 
This fact brings the advantage of allowing us to rely on
other analysis, such as the range analysis, to check if it is safe to state that
the possible values of an instruction will always be within the range of
the declared type of the instruction. 
If we can state this, it means that it is impossible for an overflow to 
occur in the analyzed instruction, because the range analysis is a conservative analysis; 
in this case, we don't need to insert dynamic checks, and the instrumented program
will be smaller and run faster than it would run without this prunning.
%How our instrumentation works
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after that instructions. 
The LLVM IR has five instructions that may lead to an overflow: 
ADD, SUB, MUL, SHL, and TRUNC.

@@

\noindent
\textbf{Precision.}
Our implementation of range analysis is remarkably precise, considering its
runtime.
Lakhdar {\em et al.}'s relational analysis~\cite{Lakhdar11}, for instance, takes
about 25 minutes to go over a program with almost 900 basic blocks.
We analyze programs of similar size in less than one second.
We do not claim our approach is as precise as such algorithms, even though we
are able to find exact bounds to 4/5 of the examples presented
in~\cite{Lakhdar11}.
On the contrary, we present a compromise between precision and speed
that scales to very large programs.
Nevertheless, our results are far from being trivial.
We have implemented a dynamic profiler that measures, for each variable,
its upper and lower limits, given an execution of the target program.
Figure~\ref{fig:precision} compares our results with those measured
dynamically for the Stanford benchmark, which is publicly
available in the LLVM test suite.
We chose Stanford because these benchmarks do not read data from external
files; hence, imprecisions are due to library functions that we cannot
analyze.

@@

\noindent
\paragraph{Dynamic Detection of Integer Overflows: }
Brumley {\em et al.}~\cite{Brumley07} have developed a tool, RICH, to
dynamically detect integer overflows.
Their goal is to secure C programs against integer overflow based
vulnerabilities.
The author's approach consists in instrumenting every integer operation that
might cause an overflow, underflow, or data loss.
The main result of Brumley {\em et al.} is the verification that instrumenting
programs to guard them against integer overflows does not compromise their
performance significantly: the average slowdown across four large applications
is 5\%.
RICH, Brumley {\em et al}'s tool, uses specific features of the x86 architecture
to reduce the instrumentation overhead.
Chinchani {\em et al.}~\cite{Chinchani04} follow a similar approach.
In this work, the authors describe each arithmetic operation formally, and then
use characteristics of the computer architecture to detect overflows at
runtime.
% TODO bibtex: cite Archerr: Runtime environment driven program safety.
Contrary to these previous work, we instrument programs at LLVM's intermediate
representation level, which is machine independent.
Nevertheless, the performance of the programs that we instrument is on par with
Brumley's.

Similar to Brumley {\em et al.}, Dietz {\em et al.}~\cite{Dietz12} have 
implemented a tool, IOC, that instruments C/C++ programs to detect integer
overflows.
They approach the problem of detecting integer overflows from a software
engineering point-of-view; hence, performance is not a concern.
The authors have used IOC to perform a lengthy study about the occurrences
of overflows in real-world programs, and have found that these events are very
common.
It is possible to implement a dynamic analysis without instrumenting the
target program.
In this case, developers must use some form of code emulation.
Chen {\em et al.}~\cite{Chen09}, for instance, uses a modified
Valgrind~\cite{Nethercote07} virtual machine to detect integer overflows.
The main drawback of emulation is performance: Chen {\em et al.} report a
50X slowdown.
% TODO: cite "Brick: A  binary tool for run-time detecting and locating
% integer-based vulnerability.
The main difference between all these previous works and ours is the fact that
we focus on generating less instrumentation, an endeavor that we accomplish via
static analysis.

\paragraph{Static Detection of Integer Overflows: }
\noindent
Zhang {\em et al.}~\cite{Zhang09} have designed a static analysis to sanitize
programs against integer overflow based vulnerabilities.
They instrument integer operations in paths from a source to a sink.
In Zhang {\em et al.}'s context, sources are functions that read values from
users, and sinks are memory allocation operations.
Thus, contrary to our work, Zhang {\em et al.}'s only need to instrument
about 10\% of the integer operations in the program.
However, they do not use any form of range analysis to limit the number of
checks inserted in the transformed code.
Wang {\em et a.}~\cite{Wang09} have implemented a tool, IntScope, that combines symbolic execution and taint analysis to detect integer overflow vulnerabilities. The authors have been able to use this tool to successfully identify many vulnerabilities in industrial quality software. Our work, and Wang {\em et a.}'s work are essentially different: they use symbolic execution, whereas we rely on
range analysis.
Furthermore, contrary to them, we do not try to detected statically the
possibility of an integer overflow happening.
Contrary to us, they do not transform the program to prevent or detect such event dynamically.
Still in the field of symbolic execution, Molnar
{\em et al.}~\cite{Molnar09} have implemented a tool, SmartFuzz, that analyzes
Linux x86 binaries to find integer overflow bugs.
They prove the existence of bugs by generating test cases for them.
